---
title: "Introduction to R"
author: "Nicholas"
date: "5/10/2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
## Conditionals and Control Flow
```{r}
# Comparison of logicals
TRUE == FALSE

# Comparison of numerics
-6*14 != 17-101

# Comparison of character strings
"useR" == "user"

# Compare a logical with a numeric
TRUE == 1

# Comparison of numerics
-6*5+2>=-10+1

# Comparison of character strings
"raining" <="raining dogs"

# Comparison of logicals
TRUE > FALSE

# The linkedin and facebook vectors have already been created for you
linkedin <- c(16, 9, 13, 5, 2, 17, 14)
facebook <- c(17, 7, 5, 16, 8, 13, 14)

# Popular days more than 15
linkedin > 15

# Quiet days, views less than 5
linkedin <= 5

# LinkedIn more popular than Facebook
linkedin > facebook

# The social data has been created for you
linkedin <- c(16, 9, 13, 5, 2, 17, 14)
facebook <- c(17, 7, 5, 16, 8, 13, 14)
views <- matrix(c(linkedin, facebook), nrow = 2, byrow = TRUE)

# When does views equal 13?
views == 13

# When is views less than or equal to 14?
views <=14

# The linkedin and last variable are already defined for you
linkedin <- c(16, 9, 13, 5, 2, 17, 14)
last <- tail(linkedin, 1)

# Is last under 5 or above 10?
last < 5 | last > 10

# Is last between 15 (exclusive) and 20 (inclusive)?
15 < last & last <= 20

# linkedin exceeds 10 but facebook below 10
linkedin > 10 & facebook < 10

# When were one or both visited at least 12 times?
linkedin >= 12 | facebook >= 12

# When is views between 11 (exclusive) and 14 (inclusive)?
between <- views > 11 & views <=14

# Count the number of Trues
sum(between==TRUE)

# Variables related to your last day of recordings
medium <- "LinkedIn"
num_views <- 14

# Examine the if statement for medium
if (medium == "LinkedIn") {print("Showing LinkedIn information")}

# Write the if statement for num_views
if(num_views > 15) {print("You are popular!")}

# Control structure for medium
if (medium == "LinkedIn") {print("Showing LinkedIn information")} else {print("Unknown medium")}

# Control structure for num_views
if (num_views > 15) {print("You're popular!")} else {print("Try to be more visible!")}

# Control structure for medium
if (medium == "LinkedIn") {
  print("Showing LinkedIn information")
} else if (medium == "Facebook") {
  print("Showing Facebook information")
} else {
  print("Unknown medium")
}

# Control structure for num_views
if (num_views > 15) {
  print("You're popular!")
} else if (num_views <= 15 & num_views > 10) {
print("Your number of views is average")
} else {
  print("Try to be more visible!")
}

# Variables related to your last day of recordings
li <- 15
fb <- 9

# Code the control-flow construct
if (li>15& fb>15) {
  sms <- 2 * (li + fb)
} else if (li<10&fb<10) {
  sms <- 0.5 * (li + fb)
} else {
  sms <- li + fb
}
sms
```

## Loops
```{r}
# Initialize the speed variable
speed <- 64

# Code the while loop
while (speed>30) {
  print("Slow down!")
  speed <- speed -7
}

# Print out the speed variable
speed

# Initialize the speed variable
speed <- 64

# Extend/adapt the while loop
while (speed > 30) {
  print(paste("Your speed is",speed))
  if (speed>48) {
    print("Slow down big time!")
    speed <- speed -11
  } else {
    print("Slow down!")
    speed <- speed -6
  }
}

# Print out the speed variable
speed

# Initialize the speed variable
speed <- 88

while (speed > 30) {
  print(paste("Your speed is", speed))
  
  # Break the while loop when speed exceeds 80
  if ( speed > 80) {
    break
  }
  
  if (speed > 48) {
    print("Slow down big time!")
    speed <- speed - 11
  } else {
    print("Slow down!")
    speed <- speed - 6
  }
}

# Initialize i as 1 
i <- 1

# Code the while loop
while (i <= 10) {
  print(3*i)
  if (3*i%%8==0) {
    break
  }
  i <- i + 1
}

# The linkedin vector has already been defined for you
linkedin <- c(16, 9, 13, 5, 2, 17, 14)

# For Loop version 1
for (i in linkedin) {print(i)}

# For Loop version 2
for (i in 1:length(linkedin)) {print(linkedin[i])}

# The nyc list is already specified
nyc <- list(pop = 8405837, 
            boroughs = c("Manhattan", "Bronx", "Brooklyn", "Queens", "Staten Island"), 
            capital = FALSE)

# Loop version 1
for (i in nyc){print(i)}

# Loop version 2
for (i in 1:length(nyc)){print(nyc[[i]])}

ttt <- matrix(1:9,nrow=3,ncol=3)
# define the double for loop
for (i in 1:nrow(ttt)) {
  for (j in 1:ncol(ttt)) {
    print(paste("On row", i, "and column", j, "the board contains", ttt[i,j]))
  }
}

# The linkedin vector has already been defined for you
linkedin <- c(16, 9, 13, 5, 2, 17, 14)

# Code the for loop with conditionals
for (li in linkedin) {
  if (li>10) {print("You're popular!")} 
  else {print("Be more visible!")}
  
   # Add if statement with break
  if (li > 16) {
    print("This is ridiculous, I'm outta here!")
    break
  }
  
  # Add if statement with next
  if (li < 5) {
    print("This is too embarrassing!")
    next
  }
  print(li)
}
```

## Functions
These are the extremely important functions when it comes down to R packages:
- install.packages(), which as you can expect, installs a given package.
- library() which loads packages, i.e. attaches them to the search list on your R workspace.
- search(), to look at the currently attached packages and

Consult the documentation on the mean() function: ?mean or help(mean).
Inspect the arguments of the mean() function using the args() function.
- The mean() function computes the arithmetic mean.
- Function template: my_fun <- function(arg1, arg2) {body}
- Variables that are defined inside a function are not accessible outside that function.

You can only load packages one at a time
- library(data.table)/library("data.table")
- require(rjson)
```{r}
# Defining a function with default arguments
pow_two <- function(x,print_info=TRUE) {
  y <- x ^ 2
  if (print_info == TRUE)
  {
  print(paste(x, "to the power two equals", y))}
  return(y)
}
pow_two(22,print_info=FALSE)
```
## lapply
To put it generally, lapply takes a vector or list X, and applies the function FUN to each of its members. If FUN requires additional arguments, you pass them after you've specified X and FUN (...). The output of lapply() is a list, the same length as X, where each element is the result of applying FUN on the corresponding element of X. You can use lapply() on your own functions as well.
- lapply(X, FUN, ...)
- essentially a for loop
- you can use so-called anonymous functions in R by putting it inside lapply

lapply: apply function over list or vector, outputs list
sapply: apply function over list or vector, try to simplify list to array
vapply: apply function over list or vector, explicitly specify output format
- vapply(X, FUN, FUN.VALUE)
vapply() can be considered a more robust version of sapply(), because you explicitly restrict the output of the function you want to apply. Converting your sapply() expressions in your own R scripts to vapply() expressions is therefore a good practice
```{r}
# Code from previous exercise:
pioneers <- c("GAUSS:1777", "BAYES:1702", "PASCAL:1623", "PEARSON:1857")
split <- strsplit(pioneers, split = ":")
split_low <- lapply(split, tolower)

# Write function select_first()
select_first <- function(x) {
  x[1]
}

# Apply select_first() over split_low: names
names <- lapply(split_low,select_first)
print(names)

# Write function select_second()
select_second<-function(x) {
  x[2]
}

# Apply select_second() over split_low: years
years <- lapply(split_low,select_second)
print(years)

# Transform: use anonymous function inside lapply
names <- lapply(split_low, function(x) {
  x[1]
})
print(names)

# Transform: use anonymous function inside lapply
years <- lapply(split_low, function(x) {
  x[2]
})
print(years)

# Generic select function
select_el <- function(x, index) {
  x[index]
}

# To include arguments in laaply
names <- lapply(split_low,select_el,index=1)
print(names)
years <- lapply(split_low,select_el,index=2)
print(years)
```

## Utilities

### Useful Functions
abs(): Calculate the absolute value.
sum(): Calculate the sum of all the values in a data structure.
mean(): Calculate the arithmetic mean from a vector of numerical values,
round(): Round the values to 0 decimal places by default.

### Data Utilities
seq(): Generate sequences, by specifying the from, to, and by arguments.
rep(): Replicate elements of vectors and lists.
sort(): Sort a vector in ascending order. Works on numerics, but also on character strings and logicals.
rev(): Reverse the elements in a data structures for which reversal is defined.
str(): Display the structure of any R object.
append(): Merge vectors or lists.
is.*(): Check for the class of an R object.
as.*(): Convert an R object from one class to another.
unlist(): Flatten (possibly embedded) lists to produce a vector.

### Regular Expressions
grepl(), which returns TRUE when a pattern is found in the corresponding character string.
grep(), which returns a vector of indices of the character strings that contains the pattern

- You can use the caret, ^, and the dollar sign, $ to match the content located in the start and end of a string, respectively. 
- which matches any character (.) zero or more times (*). Both the dot and the asterisk are metacharacters. You can use them to match any character between the at-sign and the ".edu" portion of an email address.
- \\.edu$, to match the ".edu" part of the email at the end of the string. The \\ part escapes the dot: it tells R that you want to use the . as an actual character.
- \\s: Match a space. The "s" is normally a character, escaping it (\\) makes it a metacharacter.
- [0-9]+: Match the numbers 0 to 9, at least once (+).
- ([0-9]+): The parentheses are used to make parts of the matching string available to define the replacement. 
- The \\1 in the replacement argument of sub() gets set to the string that is captured by the regular expression [0-9]+.

While grep() and grepl() were used to simply check whether a regular expression could be matched with a character vector, sub() and gsub() take it one step further: you can specify a replacement argument. sub() only replaces the first match, whereas gsub() replaces all matches.

### Dates & Times
In R, dates are represented by Date objects, while times are represented by POSIXct objects. Under the hood, however, these dates and times are simple numerical values. 
- Date objects store the number of days since the 1st of January in 1970. 
- POSIXct objects on the other hand, store the number of seconds since the 1st of January in 1970.
- Of course, it's also possible to create dates and times before 1970; the corresponding numerical values are simply negative in this case.

To create a Date object from a simple character string in R, you can use the as.Date() function. The character string has to obey a format that can be defined using a set of symbols (the examples correspond to 13 January, 1982):

%Y: 4-digit year (1982)
%y: 2-digit year (82)
%m: 2-digit month (01)
%d: 2-digit day of the month (13)
%A: weekday (Wednesday)
%a: abbreviated weekday (Wed)
%B: month (January)
%b: abbreviated month (Jan)
The following R commands will all create the same Date object for the 13th day in January of 1982:

as.Date("1982-01-13")
as.Date("Jan-13-82", format = "%b-%d-%y")
as.Date("13 January, 1982", format = "%d %B, %Y")

Notice that the first line here did not need a format argument, because by default R matches your character string to the formats "%Y-%m-%d" or "%Y/%m/%d"

In addition to creating dates, you can also convert dates to character strings that use a different date notation. For this, you use the format() function.

For a full list of conversion symbols, consult the strptime documentation in the console:
%H: hours as a decimal number (00-23)
%I: hours as a decimal number (01-12)
%M: minutes as a decimal number
%S: seconds as a decimal number
%T: shorthand notation for the typical format %H:%M:%S
%p: AM/PM indicator

as.POSIXct() uses a default format to match character strings.

Both Date and POSIXct R objects are represented by simple numerical values under the hood. This makes calculation with time and date objects very straightforward: R performs the calculations using the underlying numerical values, and then converts the result back to human-readable time information again.

You can increment and decrement Date objects, or do actual calculations with them

Calculations using POSIXct objects are completely analogous to those using Date objects.
```{r}
awards <- c("Won 1 Oscar.",
  "Won 1 Oscar. Another 9 wins & 24 nominations.",
  "1 win and 2 nominations.",
  "2 wins & 3 nominations.",
  "Nominated for 2 Golden Globes. 1 more win & 2 nominations.",
  "4 wins & 1 nomination.")

sub(".*\\s([0-9]+)\\snomination.*$", "1", awards)

today <- Sys.Date()
format(Sys.Date(), format = "%d %B, %Y")
format(Sys.Date(), format = "Today is a %A!")

today <- Sys.Date()
today + 1
today - 1

as.Date("2015-03-12") - as.Date("2015-02-27")

now <- Sys.time()
now + 3600          # add an hour
now - 3600 * 24     # subtract a day

birth <- as.POSIXct("1879-03-14 14:37:23")
death <- as.POSIXct("1955-04-18 03:47:12")
einstein <- death - birth
einstein
```
