---
title: "Introduction to R"
author: "Nicholas"
date: "5/10/2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
## Data Wrangling

Observations are stored in rows
Variables are stored in columns

You'll often need to use the pipe operator (%>%) to combine multiple dplyr verbs in a row. 
dataset %>% 

gapminder %>%
  filter(year == 1957) %>%
  arrange(desc(pop))

The filter verb extracts particular observations based on a condition. You can also use the filter() verb to set two conditions, which could retrieve a single observation.

filter(year==1957)  
filter(year==2002,country=="China")
gapminder_1952 <- gapminder %>% filter (year==1952)

You use arrange() to sort observations in ascending or descending order of a particular variable.

arrange(gapminder,lifeExp)
arrange(gapminder,lifeExp,descending=TRUE)

You can use the mutate() verb to change this column, or to create a new column that's calculated this way.
Use mutate to change lifeExp to be in months
gapminder %>%
  mutate(lifeExp = lifeExp * 12)

Use mutate to create a new column called lifeExpMonths
gapminder %>%
  mutate(lifeExpMonths = lifeExp * 12)

```{r}
library(dplyr)

```

## GGplot

### Scatter Plots
- Creating a graph: ggplot(gapminder_1952, aes(x = pop, y = gdpPercap)) + geom_point()
- X axis with log scale: ggplot(gapminder_1952, aes(x = pop, y = lifeExp)) + geom_point() + scale_x_log10()
- Different colour for category: ggplot(gapminder_1952, aes(x = pop, y = lifeExp, color = continent)) + geom_point() + scale_x_log10()
- Different sized plots: ggplot(gapminder_1952, aes(x = pop, y = lifeExp, color = continent, size = gdpPercap)) +
  geom_point() +
  scale_x_log10()
- Faceting (Subplots): ggplot(gapminder_1952, aes(x = pop, y = lifeExp, color = continent, size = gdpPercap)) +
  geom_point() +
  scale_x_log10() +
  facet_wrap(~continent)
- ggplot(by_year, aes(x=year,y=medianLifeExp),expand_limits(y=0)) + geom_point()

### Line Plots
- ggplot(by_year,aes(x=year,y=medianGdpPercap),expand_limits(y=0)) + geom_line()
  
## Grouping and Summarizing
summarize turns many rows into 1
1. medianLifeExp <- gapminder %>% summarize(median(lifeExp))
2. gapminder %>%
    filter(year == 1957) %>%
    summarize(medianLifeExp = median(lifeExp))
3. gapminder %>%
    filter(year == 1957) %>%
    summarize(medianLifeExp = median(lifeExp),
            maxGdpPercap = max(gdpPercap))
group_by before summarize turns groups into 1 row each
1. gapminder %>%
    group_by(year) %>%
    summarize(medianLifeExp = median(lifeExp), maxGdpPercap = max(gdpPercap))
2. gapminder %>%
    filter(year==1957) %>%
    group_by(continent) %>%
    summarize(medianLifeExp = median(lifeExp), maxGdpPercap = max(gdpPercap))
3. gapminder %>%
    group_by(continent, year) %>%
    summarize(medianLifeExp = median(lifeExp), maxGdpPercap = max(gdpPercap))